% chapters/ch01.tex
% Lecture 15 — CUDA Programming
% COL380: Introduction to Parallel and Distributed Programming

\chapter{CUDA Programming}
\label{ch:cuda}

% ─────────────────────────────────────────────────────────────────────────────
\section{GPU Model of Computation}
\label{sec:gpu-model}

A CUDA program runs partly on the \ppterm{host} (CPU) and partly on the
\ppterm{device} (GPU). The host orchestrates the entire computation: it
allocates memory on both sides, transfers data to the GPU, launches work, and
retrieves results. The device executes the compute-intensive portions of the
program, which are expressed as special functions called \ppterm{kernels}.

\begin{ppkey}
The host sends both data \emph{and} code to the device. Even device-side
heap memory must be allocated through host-level system calls---the host OS
issues the necessary driver calls on the program's behalf.
\end{ppkey}

\subsection{Host and Device}
\label{ssec:host-device}

The host machine is a conventional CPU system. Within a CUDA program, the
programmer writes host code (ordinary C/C++ that runs on the CPU) and device
code (kernel functions that run on the GPU) in the same source file, saved
with the \texttt{.cu} extension. The NVIDIA compiler \texttt{nvcc} separates
the two: device code is compiled for the GPU, and host code is passed through
to the system C++ compiler (typically GCC).

\subsection{The GPU Execution Hierarchy}
\label{ssec:execution-hierarchy}

The logical execution hierarchy inside a GPU, from finest to coarsest, is:

\begin{itemize}
  \item \ppterm{Thread} --- the smallest unit of execution. Each thread runs
    an instance of the kernel with its own register state.
  \item \ppterm{Warp} --- 32 threads that execute in lockstep on the same
    functional units. The warp is the hardware scheduling unit.
  \item \ppterm{Thread Block (TB)} --- a programmer-visible group of threads
    mapped to a single Streaming Multiprocessor (SM). Threads within a block
    can communicate through fast on-chip shared memory.
  \item \ppterm{TPC (Texture Processing Cluster)} --- a grouping of SMs.
  \item \ppterm{GPC (Graphics Processing Cluster)} --- a grouping of TPCs.
    A kernel is the unit of computation assigned to a GPC.
  \item \ppterm{GPU} --- the full device, comprising all GPCs.
\end{itemize}

\begin{ppnote}
Tensor Cores (four per SM) are not assigned per-thread. The SM warp scheduler
identifies warps whose instructions require tensor arithmetic and routes them
to an available Tensor Core. A warp of 32 threads is the basic execution unit
for a Tensor Core operation.
\end{ppnote}

\subsection{Parallel Kernel Execution and Streams}
\label{ssec:streams}

Multiple kernels can execute concurrently on the same GPU provided there are
unallocated SMs. Concurrent launches are achieved through \ppterm{CUDA Streams}:
an ordered queue of GPU operations in which operations across different streams
may overlap. By default every kernel launch uses the null stream, which
serializes launches; explicit streams are required for true concurrency.


% ─────────────────────────────────────────────────────────────────────────────
\section{Kernels: Definition, Structure, and the \texttt{\_\_global\_\_} Qualifier}
\label{sec:kernels}


\subsection{Definition of a Kernel}
A \textbf{kernel} is the fundamental unit of computation on a GPU. It is a function executed on the \emph{device} (GPU) and invoked from the \emph{host} (CPU). At the hardware level, a kernel represents the unit of work assigned to a \textbf{GPC (Graphics Processing Cluster)}, and is executed in parallel across many threads.

\subsection{Defining a Kernel with \_\_global\_\_}

\begin{ppprogram}{Example CUDA Kernel}{prog:cuda-vector-add}
\begin{lstlisting}[language=C++]
__global__ void vector_add(const float *A,
                           const float *B,
                           float *C,
                           int N)
{
    int i = blockIdx.x * blockDim.x + threadIdx.x;

    if (i < N) {
        C[i] = A[i] + B[i];
    }
}
\end{lstlisting}
\end{ppprogram}

\begin{ppnote}
The \verb|__global__| qualifier has precise meaning:
\begin{itemize}
    \item It is a \textbf{compiler directive} interpreted by \texttt{nvcc}.
    \item It instructs the compiler to compile the function for execution on the device and make it callable from the host.
    \item Such functions are executed on the device but invoked from the host.
\end{itemize}
\end{ppnote}

\subsection{Kernel Syntax Rules}

\begin{ppnote}
CUDA kernels follow strict syntactic and semantic constraints:
\begin{enumerate}
    \item \textbf{Return Type:} Always \verb|void|, since many threads execute concurrently with no single return value.

    \item \textbf{Parameter Passing:}
    \begin{itemize}
        \item Passed by value (e.g., \verb|int N|)
        \item Passed as pointers (e.g., \verb|float *A|)
    \end{itemize}
    Pointer arguments must refer to \textbf{device memory}.

    \item \textbf{SPMD Execution Model:}
    \begin{itemize}
        \item All threads execute the same program.
        \item Each thread operates on different data using its index.
    \end{itemize}

\begin{lstlisting}[language=C++]
int i = blockIdx.x * blockDim.x + threadIdx.x;
\end{lstlisting}

    \item \textbf{Boundary Checks:}
\begin{lstlisting}[language=C++]
if (i < N) {
    C[i] = A[i] + B[i];
}
\end{lstlisting}
    Missing this leads to undefined memory behavior.
\end{enumerate}
\end{ppnote}

\subsection{Kernel Invocation}

\begin{ppprogram}{Kernel Launch Syntax}{prog:cuda-launch}
\begin{lstlisting}[language=C++]
int threads_per_block = 256;
int blocks = (N + threads_per_block - 1) / threads_per_block;

vector_add<<<blocks, threads_per_block>>>(d_A, d_B, d_C, N);
cudaDeviceSynchronize();
\end{lstlisting}
\end{ppprogram}

\begin{ppnote}
\begin{itemize}
    \item \verb|<<<blocks, threads_per_block>>>| specifies grid and block size.
    \item Kernel launches are \textbf{asynchronous} by default.
    \item \verb|cudaDeviceSynchronize()| enforces completion.
\end{itemize}
\end{ppnote}

\subsection{Source File Organization}

\begin{ppnote}
\begin{itemize}
    \item Kernel and host code coexist in the same \verb|.cu| file.
    \item The CUDA compiler \texttt{nvcc} separates:
    \begin{itemize}
        \item Host code $\rightarrow$ host compiler
        \item Device code $\rightarrow$ GPU binary (PTX/SASS)
    \end{itemize}
\end{itemize}
\end{ppnote}

\subsection{Kernel Compilation and Loading}

\begin{ppnote}
\begin{itemize}
    \item Kernel binaries are \textbf{not} sent at compile time.
    \item They are transferred at \textbf{first invocation} (lazy loading).
\end{itemize}
\end{ppnote}

\begin{ppnote}
\textbf{Implications:}
\begin{itemize}
    \item First launch incurs overhead (loading / JIT compilation).
    \item Subsequent launches are faster.
\end{itemize}
\end{ppnote}

\subsection{Execution Mapping to Hardware}

\begin{ppnote}
\begin{itemize}
    \item Kernel $\rightarrow$ GPCs
    \item Thread blocks $\rightarrow$ SMs
    \item Threads $\rightarrow$ executed within SMs
\end{itemize}

A thread block executes entirely on a single SM, enabling efficient shared memory usage.
\end{ppnote}

\subsection{Common Pitfalls}

\begin{pppitfall}
\begin{itemize}
    \item Missing \verb|__global__| $\rightarrow$ compiled for host
    \item Using host pointers in kernel $\rightarrow$ invalid access
    \item Missing bounds check $\rightarrow$ memory corruption
    \item Assuming synchronous execution $\rightarrow$ race conditions
\end{itemize}
\end{pppitfall}

% ─────────────────────────────────────────────────────────────────────────────
\section{Thread Indexing and the Bounds Check}
\label{sec:thread-indexing}

\subsection{The SPMD Execution Model}
\label{ssec:spmd}

CUDA follows a \ppterm{Single Program Multiple Data} (SPMD) model: every
thread executes identical kernel code, but each thread operates on a different
data element determined by its unique thread index. The programmer writes the
computation for a \emph{single} element; the hardware replicates it across all
launched threads.

\subsection{Computing the Globally Unique Thread Index}
\label{ssec:thread-index}

Three built-in read-only variables identify a thread's position:

\begin{itemize}
  \item \ppinline{threadIdx.x} --- index within the block (resets to 0 in
    every block; \emph{not} globally unique).
  \item \ppinline{blockIdx.x} --- index of the block within the grid.
  \item \ppinline{blockDim.x} --- number of threads per block (same for all
    blocks in a launch).
\end{itemize}

The standard globally unique index formula for a one-dimensional grid is:
\[
  i \;=\; \texttt{blockIdx.x} \times \texttt{blockDim.x}
          \;+\; \texttt{threadIdx.x}
\]

For a two-dimensional grid (used in matrix kernels such as
Program~\ref{prog:tensor-cuda-naive}):
\[
  \texttt{row} = \texttt{blockIdx.y} \times \texttt{blockDim.y}
               + \texttt{threadIdx.y}, \quad
  \texttt{col} = \texttt{blockIdx.x} \times \texttt{blockDim.x}
               + \texttt{threadIdx.x}
\]

\subsection{Why the Bounds Check \texttt{if (i < N)} Is Required}
\label{ssec:bounds-check}

The number of blocks must be an integer, so block count is computed using
\emph{ceiling} division:
\[
  \text{blocks}
  = \left\lceil \frac{N}{\text{threadsPerBlock}} \right\rceil
  = \frac{N + \text{threadsPerBlock} - 1}{\text{threadsPerBlock}}
\]

Ceiling division may create surplus threads in the last block whose index
$i \geq N$. Without the guard \ppinline{if (i < N)}, those threads perform
silent out-of-bounds memory writes---no segmentation fault is raised.

\begin{pppitfall}
Omitting \ppinline{if (i < N)} when using ceiling block-count arithmetic
causes silent out-of-bounds writes to device memory. The corruption produces
no runtime error, making it one of the most common and hardest-to-diagnose
CUDA bugs.
\end{pppitfall}

\begin{ppexample}
$N = 34$, \ppinline{threadsPerBlock = 16}.
$\lceil 34/16 \rceil = 3$ blocks $\Rightarrow$ 48 threads total.
Threads $34, 35, \ldots, 47$ (14 threads in the last block) have no
corresponding array element and must be suppressed by the bounds check.
\end{ppexample}

Program~\ref{prog:vec-add-bounded} shows the corrected, production-style
vector addition kernel that incorporates both multi-block indexing and the
bounds check.

\begin{ppprogram}{Vector addition kernel with bounds check}{prog:vec-add-bounded}
\begin{lstlisting}[language=C++]
__global__ void vector_add(const float *A, const float *B,
                           float *C, int N)
{
    int i = blockIdx.x * blockDim.x + threadIdx.x;
    if (i < N)
        C[i] = A[i] + B[i];
}
\end{lstlisting}
\end{ppprogram}

% ─────────────────────────────────────────────────────────────────────────────

\section{Device Qualifiers: \texttt{\_\_global\_\_}, \texttt{\_\_device\_\_}, \texttt{\_\_host\_\_}}
\label{sec:device-qualifiers}

In the CUDA programming model, the programmer is responsible for explicitly partitioning code between the host and the device. This is managed through \ppterm{execution space qualifiers}, which act as directives for the \texttt{nvcc} compiler during the cross-compilation process. These qualifiers dictate both the execution site—where the code runs—and the invocation site—where the call originates. 

Because both host and device portions of a program are eventually packaged into a single common executable, these qualifiers ensure the runtime environment can correctly load and trigger device code onto the GPU as required.

\subsection{The \texttt{\_\_global\_\_} Qualifier: Kernels}

The \ppterm{\_\_global\_\_} qualifier defines a \textbf{kernel}, which serves as the primary entry point for parallel execution on the GPU. While a kernel executes on the device, it is uniquely characterized by its invocation from the host code.

\begin{itemize}
    \item \textbf{Asynchronous RPC Model:} Architecturally, a kernel launch functions as an asynchronous Remote Procedure Call (RPC). The host thread triggers the task across the high-latency PCIe bus and immediately proceeds to its next instruction without waiting for the GPU to finish.
    \item \textbf{Return Type Constraints:} Kernels are strictly required to have a \ppinline{void} return type. This is because they are executed across a massive grid of threads simultaneously, meaning there is no single return value to provide back to the host caller.
    \item \textbf{Execution Mapping:} At the hardware level, a kernel represents the unit of work assigned to a Graphics Processing Cluster (GPC).
\end{itemize}

\subsection{The \texttt{\_\_device\_\_} Qualifier: GPU Helpers}

Internal logic and mathematical utilities intended for use within a kernel are defined using the \ppterm{\_\_device\_\_} qualifier. These entities reside and execute entirely on the GPU.

\begin{itemize}
    \item \textbf{Restricted Invocation:} These functions can only be called from other device-side code, such as a \ppinline{\_\_global\_\_} kernel or another \ppinline{\_\_device\_\_} function. Attempting to call them from the host results in a compile-time error.
    \item \textbf{Zero PCIe Overhead:} Since the code and data already reside in the device's GDRAM, these calls do not cross the PCIe bus, making them highly efficient for frequent utility tasks.
    \item \textbf{Hardware Call Stack:} Despite the SIMT nature of GPUs, the call stack for \ppinline{\_\_device\_\_} functions mirrors traditional CPU behavior, maintaining its own program counter, stack pointer, and return addresses.
    \item \textbf{Compiler Optimizations:} Because the target architecture is known at compile time, \texttt{nvcc} applies aggressive GPU-specific optimizations, including specialized register allocation and instruction selection.
\end{itemize}

\subsection{The \texttt{\_\_host\_\_} Qualifier: CPU Logic}

Code intended for standard execution on the CPU is marked with the \ppterm{\_\_host\_\_} qualifier. This handles the orchestration of the parallel application, including memory allocation and data transmission.

\begin{itemize}
    \item \textbf{Implicit Default:} This is the default status for any function; any code defined without a CUDA-specific prefix is treated as \ppinline{\_\_host\_\_} by the compiler.
    \item \textbf{Native Translation:} During compilation, \texttt{nvcc} identifies host sections and leverages the system’s native C++ compiler (such as GCC) for translation into CPU binaries.
    \item \textbf{Resource Access:} Host functions have full access to Host RAM and standard system calls but cannot directly manipulate device-private memory without using specialized runtime functions like \texttt{cudaMemcpy}.
\end{itemize}

\begin{ppnote}
\textbf{Combining Qualifiers and Managed Memory:} 
To facilitate code reuse, CUDA allows qualifiers to be combined. A function marked \ppinline{\_\_host\_\_ \_\_device\_\_} is compiled into two distinct binaries, allowing the same logic to serve both execution spaces. Furthermore, the \ppinline{\_\_managed\_\_} keyword can be used for variables to ensure they are visible and accessible to both the host and device contexts simultaneously.
\end{ppnote}

% ─────────────────────────────────────────────────────────────────────────────
\section{Variable Qualifiers and the Memory Hierarchy}
\label{sec:var-qualifiers}

Variables in CUDA carry qualifiers that determine their physical storage
location, access latency, and lifetime. The programmer---not the hardware---is
responsible for choosing the right memory space for each data structure.

\subsection{\texttt{\_\_device\_\_} --- Global Device Memory (GDRAM)}
\label{ssec:var-device}

\begin{lstlisting}[language=C++]
__device__ float d_buffer[1024];
\end{lstlisting}

Resides off-chip in GDRAM. Scope: all threads, all blocks. Lifetime: entire
application. Latency: high (${\sim}200$--$800$ cycles uncached). Capacity:
several GB on a modern GPU. Dynamically-sized arrays are allocated via
\ppinline{cudaMalloc} from host code.

\subsection{\texttt{\_\_constant\_\_} --- Constant Memory}
\label{ssec:var-constant}

\begin{lstlisting}[language=C++]
__constant__ float coeffs[64];
\end{lstlisting}

Resides off-chip in a dedicated constant memory region backed by a dedicated
cache. Read-only from device code; the host writes to it via
\ppinline{cudaMemcpyToSymbol}. Scope: all threads. Lifetime: entire
application. Capacity: 64\,KB on most architectures.

Latency is low when all threads in a warp read the \emph{same} address
(hardware broadcasts one fetch to all 32 threads). When threads read
\emph{different} addresses the requests are serialized, eliminating the
latency advantage.

\begin{ppnote}
Constant memory is best suited for read-only values that are uniform across
threads: filter coefficients, physical constants, small lookup tables. It is
not a substitute for GDRAM when threads access distinct addresses.
\end{ppnote}

\subsection{\texttt{\_\_shared\_\_} --- Shared Memory}
\label{ssec:var-shared}

\begin{lstlisting}[language=C++]
__shared__ float tile[32][32];
\end{lstlisting}

Resides on-chip in the SM's SRAM. On modern NVIDIA GPUs the L1 data cache and
shared memory occupy the same physical SRAM; the split between them is
software-configurable (dual-configurable). Scope: threads within the same
block only. Lifetime: block lifetime. Latency: very low (${\sim}1$--$4$
cycles). Capacity: small---typically 48\,KB to 164\,KB per SM, shared among
all resident blocks.

Because shared memory is scarce, the programmer must explicitly decide what
data to stage there. It is the basis for tiling optimizations (covered in the
next lecture).

\begin{pppitfall}
Shared memory is private to a block. Data written by one block is never
visible to another block. Cross-block communication must go through GDRAM.
\end{pppitfall}

\subsection{Memory Hierarchy Summary}
\label{ssec:mem-summary}

% NOTE FOR EDITOR: Insert the GPU memory hierarchy figure from lec15-CUDA.pdf
% here, showing the layered hierarchy from registers through shared/L1,
% L2 cache, GDRAM, to host DRAM, annotated with latencies and capacities.

\begin{table}[h]
\centering
\caption{GPU memory spaces and their properties.}
\label{tab:mem-hierarchy}
\begin{tabular}{lllll}
\hline
\textbf{Qualifier} & \textbf{Location} & \textbf{Latency} &
  \textbf{Scope} & \textbf{Lifetime} \\
\hline
(register)              & On-chip registers & $\sim$1 cycle         & Thread      & Thread \\
\ppinline{\_\_shared\_\_}   & On-chip SRAM      & $\sim$1--4 cycles     & Block       & Block  \\
\ppinline{\_\_constant\_\_} & Off-chip (cached) & Low (broadcast)       & All threads & App    \\
\ppinline{\_\_device\_\_}   & GDRAM             & $\sim$200--800 cycles & All threads & App    \\
\hline
\end{tabular}
\end{table}


% ─────────────────────────────────────────────────────────────────────────────
\section{Kernel Invocation, Block/Grid Sizing, and Synchronization}
\label{sec:invocation}

\subsection{The Chevron (Execution Configuration) Syntax}
\label{ssec:chevron}

A kernel is launched from host code using \ppterm{triple-chevron} notation:

\begin{lstlisting}[language=C++]
kernel_name<<<gridDim, blockDim>>>(arg1, arg2, ...);
\end{lstlisting}

\ppinline{gridDim} specifies the number of blocks (a \ppinline{dim3} value,
supporting 1D, 2D, or 3D grids). \ppinline{blockDim} specifies threads per
block. Their product gives the total number of launched threads.

\subsection{Computing the Number of Blocks}
\label{ssec:num-blocks}

The number of threads per block is programmer-specified, subject to a hardware
upper limit (e.g.\ 2048 on the NVIDIA A100; determined by the SM's register
file size). A common default is 256. The block count is computed via ceiling
division:

\begin{lstlisting}[language=C++]
int threads_per_block = 256;
int blocks = (N + threads_per_block - 1) / threads_per_block;
\end{lstlisting}

\begin{ppnote}
Hardware designers expose the threads-per-block limit as a tunable parameter
rather than fixing it, because the optimal value is workload-dependent.
Compute-bound workloads may favour fewer, larger blocks (fewer active SMs,
less inter-SM overhead); memory-bound workloads benefit from more blocks to
keep the memory pipeline busy through latency hiding.
\end{ppnote}

\subsection{\texttt{cudaDeviceSynchronize()}}
\label{ssec:device-sync}

Kernel launches are \emph{asynchronous}: the \ppinline{<<<...>>>} call
returns immediately and the host continues executing subsequent statements
while the GPU works in parallel. At the point where the host requires the
GPU's results, the programmer must insert \ppinline{cudaDeviceSynchronize()},
which blocks the host thread until all previously issued GPU work completes.

\begin{pppitfall}
Omitting \ppinline{cudaDeviceSynchronize()} before a device-to-host
\ppinline{cudaMemcpy} is a silent race condition: the copy starts while the
kernel is still running, reading partially written or stale data with no
error message.
\end{pppitfall}

\subsection{\texttt{cudaMalloc} and \texttt{cudaMemcpy}}
\label{ssec:malloc-memcpy}

Before launching a kernel, the programmer must allocate memory on the device
and copy input data across the PCIe bus. After the kernel completes, results
are copied back.

\ppinline{cudaMalloc(\&ptr, bytes)} allocates \ppinline{bytes} bytes on the
GDRAM heap and writes the device pointer to \ppinline{ptr}. It is called from
host code and is the device analogue of \ppinline{malloc}.

\ppinline{cudaMemcpy(dst, src, bytes, direction)} transfers data. The
direction flag is one of:
\begin{itemize}
  \item \ppinline{cudaMemcpyHostToDevice} (\ppterm{H2D}) --- CPU $\to$ GPU,
    before the kernel.
  \item \ppinline{cudaMemcpyDeviceToHost} (\ppterm{D2H}) --- GPU $\to$ CPU,
    after the kernel.
\end{itemize}

\ppinline{cudaMemcpy} is \emph{synchronous} with respect to the host by
default: it does not return until the transfer is complete. H2D and D2H
transfers are high-latency operations over PCIe; they eat into the time saved
by GPU parallelism and should be minimised.

Program~\ref{prog:data-transfer} shows the complete four-phase lifecycle.

\begin{ppprogram}{Complete CUDA data-transfer and invocation lifecycle}{prog:data-transfer}
\begin{lstlisting}[language=C++]
/* --- Step 1: allocate on device --- */
float *d_A, *d_B, *d_C;
CUDA_CHECK(cudaMalloc(&d_A, bytes));
CUDA_CHECK(cudaMalloc(&d_B, bytes));
CUDA_CHECK(cudaMalloc(&d_C, bytes));

/* --- Step 2: copy inputs H2D --- */
CUDA_CHECK(cudaMemcpy(d_A, h_A, bytes, cudaMemcpyHostToDevice));
CUDA_CHECK(cudaMemcpy(d_B, h_B, bytes, cudaMemcpyHostToDevice));

/* --- Step 3: launch kernel --- */
int blocks = (N + 255) / 256;
vector_add<<<blocks, 256>>>(d_A, d_B, d_C, N);
CUDA_CHECK(cudaDeviceSynchronize());

/* --- Step 4: copy result D2H --- */
CUDA_CHECK(cudaMemcpy(h_C, d_C, bytes, cudaMemcpyDeviceToHost));

/* --- Step 5: free device memory --- */
CUDA_CHECK(cudaFree(d_A)); CUDA_CHECK(cudaFree(d_B));
CUDA_CHECK(cudaFree(d_C));
\end{lstlisting}
\end{ppprogram}

\subsection{Inter-SM Memory Sharing}
\label{ssec:inter-sm}

Within a block, threads share \ppinline{\_\_shared\_\_} memory on the same SM and
synchronize with \ppinline{\_\_syncthreads()}. Across SM boundaries, the only
memory accessible to all blocks is the L2 cache and GDRAM---there is no
cross-SM shared memory. Cross-block communication therefore requires staging
data through GDRAM and, in general, a second kernel launch to ensure all
writes are visible before any reads.

\begin{ppkey}
Within a block: synchronize via \ppinline{\_\_shared\_\_} memory.
Across blocks: communicate only through L2 cache or GDRAM.
\end{ppkey}


% ─────────────────────────────────────────────────────────────────────────────
\section{Code Walkthrough: \texttt{tensor\_bench.cu}}
\label{sec:code-walkthrough}

To make everything concrete, let us walk through a real benchmark.
The file \texttt{tensor\_bench.cu} implements the tensor contraction:

\[
  C[N][M] \mathrel{+}= A[N][K] \times B[K][M], \quad N = M = K = 2048
\]
Think of this as a matrix multiplication where the result is accumulated into
$C$ rather than overwritten. With $N = M = K = 2048$ the computation requires
$2 \times 2048^3 \approx 34.4$\,GFLOP --- enough work to make the differences
between implementations clearly visible.
 
The file contains five variants. We will walk through the first three in
detail. \textbf{Variants 4 (OMP tiled) and 5 (CUDA tiled with shared memory)
are intentionally skipped here}---shared-memory tiling is the subject of the
next lecture.
 
\begin{ppnote}
The five variants in order are:
\begin{enumerate}
  \item Sequential --- plain triple-loop, single core.
  \item OMP flat --- \ppinline{collapse(2)} + SIMD.
  \item CUDA naive --- one thread per output element, global memory only.
  \item OMP tiled --- $32{\times}32$ CPU cache tiles (next lecture).
  \item CUDA tiled --- shared-memory tiling (next lecture).
\end{enumerate}
\end{ppnote}


%─────────────────────────────────────────────────────────────────────────────
\subsection{Compiling the Benchmark}
\label{ssec:compile}
 
Because the file mixes CUDA device code and OpenMP host code, it needs both
the NVIDIA compiler and the system C compiler:
 
\begin{lstlisting}[language=bash]
nvcc -O3 -Xcompiler "-O3 -march=native -fopenmp" -lm \
     -o tensor_bench tensor_bench.cu -lgomp
\end{lstlisting}
 
Let us read each flag:
 
\begin{itemize}
  \item \ppinline{nvcc} --- the NVIDIA compiler driver. It splits the
    \texttt{.cu} file: device code goes to the GPU compiler, host code goes
    to GCC.
  \item \ppinline{-O3} --- tells \ppinline{nvcc} to optimise the device
    (GPU) code at the highest level.
  \item \ppinline{-Xcompiler "-O3 -march=native -fopenmp"} --- passes these
    three flags \emph{directly to GCC} for the host code. \ppinline{-O3}
    and \ppinline{-march=native} give maximum CPU optimisation;
    \ppinline{-fopenmp} tells GCC to recognise \ppinline{\#pragma omp}
    directives.
  \item \ppinline{-lm} --- links the standard C math library (used by
    \ppinline{fabsf}, \ppinline{sqrtf}, etc.).
  \item \ppinline{-lgomp} --- links the OpenMP \emph{runtime} library at
    link time. Note the distinction: \ppinline{-fopenmp} is a compile-time
    flag that handles pragma parsing; \ppinline{-lgomp} attaches the runtime
    support that actually creates and schedules threads.
\end{itemize}
 
\begin{ppnote}
\ppinline{-fopenmp} and \ppinline{-lgomp} do different jobs and both are
required. Omitting \ppinline{-fopenmp} means the \ppinline{\#pragma omp}
directives are silently ignored (the code compiles but runs single-threaded).
Omitting \ppinline{-lgomp} produces an undefined-symbol linker error.
\end{ppnote}


\subsection{Variant 1 --- Sequential (\texttt{tensor\_sequential})}
\label{ssec:code-seq}

\begin{ppprogram}{Tensor contraction --- sequential}{prog:tensor-seq}
\begin{lstlisting}[language=C++]
static void tensor_sequential(
    const float *A, const float *B, float *C, int n, int m, int k)
{
    for (int i = 0; i < n; i++)
        for (int p = 0; p < k; p++) {
            float a_ip = A[i*k + p];      /* hoisted scalar load */
            for (int j = 0; j < m; j++)
                C[i*m + j] += a_ip * B[p*m + j];
        }
}
\end{lstlisting}
\end{ppprogram}

The loop is ordered $i$-$p$-$j$ (ikj order). Hoisting \ppinline{A[i*k+p]}
into the scalar \ppinline{a\_ip} avoids one redundant GDRAM load per $j$
iteration. Both \ppinline{B[p*m+j]} and \ppinline{C[i*m+j]} are accessed with
unit stride, which is cache-friendly on the CPU. This variant produces the
reference output used to verify all other variants.


\subsection{Variant 2 --- OMP Flat (\texttt{tensor\_omp\_flat})}
\label{ssec:code-omp}

\begin{ppprogram}{Tensor contraction --- OMP flat}{prog:tensor-omp}
\begin{lstlisting}[language=C++]
static void tensor_omp_flat(
    const float *A, const float *B, float *C, int n, int m, int k)
{
    #pragma omp parallel for schedule(static) collapse(2)
    for (int i = 0; i < n; i++)
        for (int p = 0; p < k; p++) {
            float a_ip = A[i*k + p];
            #pragma omp simd
            for (int j = 0; j < m; j++)
                C[i*m + j] += a_ip * B[p*m + j];
        }
}
\end{lstlisting}
\end{ppprogram}

\ppinline{collapse(2)} flattens the $(i,p)$ loop nest into a single iteration
space of $n \times k = 2048^2 \approx 4.2\,\text{M}$ iterations, giving the
OpenMP runtime far more work to distribute evenly across the 40 threads of the
server used in the demonstration---compared with only $n = 2048$ iterations if
just the outer loop were parallelized. The innermost $j$-loop is vectorized
with \ppinline{\#pragma omp simd}. The matrices are stored as flattened
one-dimensional arrays so that all accesses are stride-1 in memory.


\subsection{Variant 3 --- CUDA Naive (\texttt{kernel\_naive})}
\label{ssec:code-cuda-naive}

\begin{ppprogram}{Tensor contraction --- CUDA naive kernel}{prog:tensor-cuda-naive}
\begin{lstlisting}[language=C++]
__global__ void kernel_naive(
    const float * __restrict__ A,
    const float * __restrict__ B,
          float * __restrict__ C,
    int n, int m, int k)
{
    int row = blockIdx.y * blockDim.y + threadIdx.y;
    int col = blockIdx.x * blockDim.x + threadIdx.x;

    if (row >= n || col >= m) return;       /* 2-D bounds check */

    float acc = 0.f;
    for (int p = 0; p < k; p++)
        acc += A[row*k + p] * B[p*m + col];

    C[row*m + col] += acc;
}
\end{lstlisting}
\end{ppprogram}

The host launches this kernel with a 2D grid:

\begin{ppprogram}{CUDA naive --- host launch with CUDA Event timing}{prog:tensor-naive-host}
\begin{lstlisting}[language=C++]
dim3 block_naive(16, 16);                          /* 256 threads/block */
dim3 grid_naive((M + 15) / 16, (N + 15) / 16);   /* ceil in each dim  */

cudaEvent_t ev0, ev1;
CUDA_CHECK(cudaEventCreate(&ev0));
CUDA_CHECK(cudaEventCreate(&ev1));

CUDA_CHECK(cudaMemset(dC, 0, SZ * sizeof(float)));
CUDA_CHECK(cudaEventRecord(ev0));
kernel_naive<<<grid_naive, block_naive>>>(dA, dB, dC, N, M, K);
CUDA_CHECK(cudaEventRecord(ev1));
CUDA_CHECK(cudaEventSynchronize(ev1));

float ms_kernel;
CUDA_CHECK(cudaEventElapsedTime(&ms_kernel, ev0, ev1));
\end{lstlisting}
\end{ppprogram}

Key points about this kernel:

\begin{description}
  \item[2D thread indexing.] \ppinline{row} and \ppinline{col} extend the 1D
    index formula to two dimensions, covering the full $N \times M$ output
    matrix.

  \item[2D bounds check.] \ppinline{if (row >= n || col >= m) return}
    suppresses surplus threads in both dimensions simultaneously.

  \item[\ppinline{\_\_restrict\_\_}.] Promises the compiler that no two pointers
    alias the same memory, enabling load/store reordering and prefetching
    optimizations.

  \item[CUDA Events for kernel-only timing.] \ppinline{cudaEventRecord}
    inserts a GPU-side timestamp. \ppinline{cudaEventElapsedTime} returns the
    GPU-measured interval in milliseconds---pure kernel execution time,
    excluding H2D and D2H PCIe transfer overhead.

  \item[DRAM pressure.] Every thread independently reads $K$ elements of $A$
    and $K$ elements of $B$ from GDRAM. Total DRAM traffic is $O(NMK)$
    reads---each element of $A$ and $B$ is re-read by many threads with no
    reuse. This is the fundamental inefficiency that shared-memory tiling
    addresses in a future lecture.
\end{description}

\subsection{Observed Results and Transfer Overhead}
\label{ssec:bench-results}

Running on a server with 40 CPU cores and an NVIDIA GPU, the benchmark
reported the following approximate results for $N = M = K = 2048$:

\begin{itemize}
  \item Sequential: ${\approx}1800$\,ms
  \item OMP flat (40 threads): ${\approx}90$\,ms \quad (${\sim}20\times$ over sequential)
  \item CUDA naive (kernel only): ${\approx}9$\,ms \quad (${\sim}200\times$ over sequential,
    ${\sim}10\times$ over OMP)
\end{itemize}

\begin{ppnote}
The H2D transfer (matrices $A$ and $B$, ${\approx}32$\,MB) and D2H transfer
(matrix $C$, ${\approx}16$\,MB) together took approximately $10$--$11$\,ms
over PCIe---comparable to the ${\approx}9$\,ms kernel execution time itself.
This illustrates a fundamental trade-off: the kernel-only speedup is
${\sim}200\times$, but the end-to-end speedup (including transfers) is
significantly lower. PCIe transfers must be minimised, and GPU offload is only
worthwhile when computation time dominates transfer time.
\end{ppnote}


\subsection{Benchmarking Infrastructure}
\label{ssec:code-bench}

The benchmark runs each variant \ppinline{RUNS = 5} times and reports the
median elapsed time to reduce OS scheduling jitter and one-time GPU warm-up
effects. Correctness is verified by computing the maximum absolute error
between each variant's output and the sequential reference:

\begin{lstlisting}[language=C++]
static float max_err(const float *ref, const float *cmp, int size) {
    float e = 0.f;
    for (int i = 0; i < size; i++) {
        float d = fabsf(ref[i] - cmp[i]);
        if (d > e) e = d;
    }
    return e;
}
\end{lstlisting}

An error below $5 \times 10^{-3}$ is reported as \texttt{OK}. Small non-zero
errors are expected between parallel variants and the sequential reference
because floating-point addition is non-associative and different execution
orders produce slightly different rounding.


%─────────────────────────────────────────────────────────────────────────────
\section{Exercises}
\label{sec:exercises}

% ── Exercise 1 ───────────────────────────────────────────────────────────────
\begin{ppexercises}
\ppexercise{%
  With $N = 1000$ and \ppinline{threadsPerBlock = 256}, how many blocks are
  launched? How many threads in the last block have no work to do? Verify
  your answer by tracing the index formula for the last block.}
\end{ppexercises}

\begin{ppnote}
\textbf{Solution.}
Using the ceiling formula:
\[
  \text{blocks}
  = \left\lceil \frac{N}{\texttt{threadsPerBlock}} \right\rceil
  = \left\lceil \frac{1000}{256} \right\rceil
  = 4
\]
Total threads launched: $4 \times 256 = 1024$.
Idle threads in the last block: $1024 - 1000 = \mathbf{24}$.

\textit{Verification.} The last block has \ppinline{blockIdx.x = 3}. Its
threads compute:
\[
  i = 3 \times 256 + \texttt{threadIdx.x}, \quad
  \texttt{threadIdx.x} \in [0,\,255]
  \;\Longrightarrow\;
  i \in [768,\,1023]
\]
Elements $[768, 999]$ (232 threads) are within bounds and do useful work.
Elements $[1000, 1023]$ (24 threads) have $i \geq N$ and are suppressed by
\ppinline{if (i < N)}. Without that guard, these threads would write beyond
the end of the allocated array with no runtime error.
\end{ppnote}

% ── Exercise 2 ───────────────────────────────────────────────────────────────
\begin{ppexercises}
\ppexercise{%
  Kernel launches in CUDA are asynchronous. Explain what this means for the
  host program, and why \ppinline{cudaDeviceSynchronize()} must be called
  before copying results back to the host.}
\end{ppexercises}

\begin{ppnote}
\textbf{Solution.}
When the host calls \ppinline{kernel<<<grid, block>>>(...)} the call returns
immediately --- the host continues executing subsequent statements while the
GPU runs the kernel in parallel. The host has no implicit knowledge of when
the GPU has finished.

\ppinline{cudaDeviceSynchronize()} is a blocking call that pauses the CPU
thread until all previously issued GPU work in the current context is
complete. It must be called before a device-to-host \ppinline{cudaMemcpy}
because that transfer is also asynchronous with respect to any in-flight
kernel. Without the synchronisation barrier, the copy may begin while the
kernel is still writing to device memory, producing partially written or
stale results with no error message.
\end{ppnote}

% ── Exercise 3 ───────────────────────────────────────────────────────────────
\begin{ppexercises}
\ppexercise{%
  A student removes the guard \ppinline{if (i < N)} from a vector-addition
  kernel that uses ceiling block-count arithmetic. The program gives correct
  output when $N = 256$ (with \ppinline{threadsPerBlock = 256}) but wrong
  output when $N = 300$. Explain both observations.}
\end{ppexercises}

\begin{ppnote}
\textbf{Solution.}

\textit{Correct for $N = 256$.}
Ceiling division gives $\lceil 256/256 \rceil = 1$ block of exactly 256
threads. Every thread computes $i \in [0, 255]$, all of which are valid
indices. No surplus threads are created, so removing the guard has no
observable effect.

\textit{Wrong for $N = 300$.}
Ceiling division gives $\lceil 300/256 \rceil = 2$ blocks, launching 512
threads in total. The 212 threads with $i \in [300, 511]$ have no
corresponding array element. Without the guard, each executes
\ppinline{C[i] = A[i] + B[i]} at out-of-bounds addresses, writing into
memory adjacent to the allocated arrays. The GPU does not raise a
segmentation fault for out-of-bounds writes --- execution continues silently
and the output is corrupted.

The root cause in both cases is the same: the total thread count is always
rounded up to a whole number of blocks, so the final block almost always
contains some idle threads that must be explicitly suppressed.
\end{ppnote}

% ── Exercise 4 ───────────────────────────────────────────────────────────────
\begin{ppexercises}
\ppexercise{%
  In \ppinline{tensor\_omp\_flat}, the pragma
  \ppinline{\#pragma omp parallel for schedule(static) collapse(2)} is used.
  Why does \ppinline{collapse(2)} over the $(i, p)$ loops give better
  parallelism than collapsing only the outer $i$-loop? Give a numerical
  argument using $n = k = 2048$ and 40 threads.}
\end{ppexercises}

\begin{ppnote}
\textbf{Solution.}
Without \ppinline{collapse(2)}, only the outer $i$-loop is parallelised.
It has $n = 2048$ iterations, giving each of the 40 threads approximately
$\lfloor 2048/40 \rfloor = 51$ iterations. Within each iteration the inner
$p$-loop runs sequentially over all $k = 2048$ steps, so the parallel work
is coarse-grained and the inner serial work dominates.

With \ppinline{collapse(2)}, the $(i, p)$ space is flattened into a single
iteration space of:
\[
  n \times k = 2048 \times 2048 = 4{,}194{,}304 \text{ iterations}
\]
Each thread receives $\lfloor 4{,}194{,}304/40 \rfloor \approx 104{,}857$
iterations --- more than $2000\times$ the work compared to the $i$-only case.
The OpenMP runtime has far greater scheduling flexibility, load balance
improves across all 40 threads, and the innermost serial bottleneck is
eliminated.
\end{ppnote}

% ── Exercise 5 ───────────────────────────────────────────────────────────────
\begin{ppexercises}
\ppexercise{%
  For $N = M = K = 2048$, the CUDA naive kernel runs in approximately
  $9$\,ms (kernel only), while the H2D and D2H transfers together take
  approximately $11$\,ms. The sequential baseline takes $1800$\,ms. Compute
  the kernel-only speedup and the end-to-end speedup. What does the
  difference between these two figures tell you about GPU offload?}
\end{ppexercises}

\textbf{Solution.}

\textit{Kernel-only speedup:}
\[
  S_{\text{kernel}}
  = \frac{T_{\text{seq}}}{T_{\text{kernel}}}
  = \frac{1800\,\text{ms}}{9\,\text{ms}}
  = 200\times
\]

\textit{End-to-end time} (kernel $+$ transfers):
\[
  T_{\text{e2e}} = 9\,\text{ms} + 11\,\text{ms} = 20\,\text{ms}
\]

\textit{End-to-end speedup:}
\[
  S_{\text{e2e}}
  = \frac{T_{\text{seq}}}{T_{\text{e2e}}}
  = \frac{1800\,\text{ms}}{20\,\text{ms}}
  = \mathbf{90\times}
\]

The kernel-only speedup is $200\times$, but the true application-level
speedup is $90\times$ because the PCIe transfers cost nearly as much as the
kernel itself. H2D and D2H are high-latency operations that directly reduce
the net benefit of GPU parallelism. GPU offload is only worthwhile when the
computation is large enough that kernel time dominates transfer time. For
workloads that reuse the same data across many kernel launches, the transfer
cost is amortised over multiple runs, making the kernel-only speedup a more
accurate predictor of overall performance.